Comparing Power Optimizations in the Reorder Buffer

### 2011-06-08 paper/paper.tex
Paper sections for circular ROB and dynamic + dynamic optimized ROBs done.

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is commited from the ROB to the architectrual register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is 
\subsection{Latch ROB}
\subsection{Dynamic ROB}
\subsection{Distributed ROB}

@@

The instruction generator is currently capable of generating four distinct instruction streams corresponding to different instruction workloads.  The first instruction stream is marked as the default instruction stream.  It features equal percentages of each type of instruction as well as a 10% register dependence rate.  This register dependence rate was selected arbitrarily to act as a starting point for modeling data dependencies between instructions.

@@

\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
ins{\char'137}profile&default{\char'137}profile&173{\char'056}applu&191{\char'056}fma3d&183{\char'056}equake\\ \hline
loadProc&0.10&0.32&0.12&0.37\\ \hline
storeProc&0.10&0.12&0.24&0.06\\ \hline
addProc&0.10&0.01&0.15&0.08\\ \hline
subProc&0.10&0.01&0.15&0.08\\ \hline
multProc&0.10&0.01&0.15&0.08\\ \hline
divProc&0.10&0.01&0.15&0.08\\ \hline
faddProc&0.10&0.13&0.01&0.07\\ \hline
fsubProc&0.10&0.13&0.01&0.06\\ \hline
fmultProc&0.10&0.13&0.01&0.06\\ \hline
fdivProc&0.10&0.13&0.01&0.06\\ \hline
forwardingRate&0.10&0.06&0.06&0.06\\ \hline
instCount&10000&10000&10000&10000\\ \hline\end{tabular}
\caption{Benchmark Instruction Profiles showing percent rate of occurrence of each instruction type.}
\label{tab:insprofile}
\end{table*}

For the purposes of our power comparisons, each of the above instruction workloads are normalized to generate 10000 instructions.  Such a large instruction window is ideal to allow buffers to fill and behavioral characteristics to develop.  The specific values used for each workload are listed in Table ~\ref{tab:insprofile}.
\subsection{Model Implementation}
The following section details the model implementatations in our simulator.  The basic re-order buffer model we use as a base-line for comparison is a standard circular ROB model.  Section ~\ref{sec:circ} discusses its implementation.  We further model three optimizations for re-order buffers in an attempt to improve overall power usage: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch).  Each of these optimized models are built on top of the underlying circular ROB model to provide as close a behavioral and power comparison as possible.  Each of the optimizations, as well as the standard circular reorder buffer, are described in detail below.
\subsubsection{Circular ROB}
\label{sec:circ}
\subsubsection{Distributed (Cluster based) ROB}
\subsubsection{Dynamically Sized ROB}
\subsubsection{Reduced Output Port ROB (Retention Latch)}
\subsection{Power Tabulation Implementation}
\section{Results}
hi
\subsection{subsection titl}
sub section.. with table!
\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

@@

%TODO add in bib 
\begin{thebibliography}{------}
\bibitem{graphite}
	MIT Graphite Simulator
	https://github.com/mit-carbon/Graphite/wiki
\bibitem{kucuk}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Low-complexity reorder buffer architecture. In Proceedings
of the 16th international conference on Supercomputing (ICS '02). ACM, New York, NY, USA, 57-66. DOI=10.
1145/514191.514202 http://doi.acm.org/10.1145/514191.514202
\bibitem{rupnow}
Kyle Rupnow, Arun Rodrigues, Keith Underwood, and Katherine Compton. 2006. Scientific applications vs. SPEC-FP: a comparison of program behavior. In Proceedings of the 20th annual international conference on Supercomputing (ICS '06). ACM, New York, NY, USA, 66-74. DOI=10.1145/1183401.1183413 http://doi.acm.org/10.1145/1183401.1183413 
\end{thebibliography}
\end{document}

>>>>

Upon execution completion, the output value is stored in the ROB with its
corresponding entry and that entry is marked as valid or completed with a
1 bit flag.  An entry is commited from the ROB to the architectrual register
file (ARF) when that entry is the oldest (lowest PC) and its result has been
marked as valid.

@@

The instruction generator is currently capable of generating four distinct instruction streams corresponding to different instruction workloads.  The first instruction stream is marked as the default instruction stream.  It features equal percentages of each type of instruction as well as a 10\% register dependence rate.  This register dependence rate was selected arbitrarily to act as a starting point for modeling data dependencies between instructions.

@@

\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
ins{\char'137}profile&default{\char'137}profile&173{\char'056}applu&191{\char'056}fma3d&183{\char'056}equake\\ \hline
loadProc&0.10&0.32&0.12&0.37\\ \hline
storeProc&0.10&0.12&0.24&0.06\\ \hline
addProc&0.10&0.01&0.15&0.08\\ \hline
subProc&0.10&0.01&0.15&0.08\\ \hline
multProc&0.10&0.01&0.15&0.08\\ \hline
divProc&0.10&0.01&0.15&0.08\\ \hline
faddProc&0.10&0.13&0.01&0.07\\ \hline
fsubProc&0.10&0.13&0.01&0.06\\ \hline
fmultProc&0.10&0.13&0.01&0.06\\ \hline
fdivProc&0.10&0.13&0.01&0.06\\ \hline
forwardingRate&0.10&0.06&0.06&0.06\\ \hline
instCount&100000&100000&100000&100000\\ \hline\end{tabular}
\caption{Benchmark Instruction Profiles showing percent rate of occurrence of each instruction type.}
\label{tab:insprofile}
\end{table*}

For the purposes of our power comparisons, each of the above instruction workloads are normalized to generate 10000 instructions.  Such a large instruction window is ideal to allow buffers to fill and behavioral characteristics to develop.  The specific values used for each workload are listed in Table ~\ref{tab:insprofile}.
\subsection{Model Implementation}
The following section details the model implementatations in our simulator.  The basic re-order buffer model we use as a base-line for comparison is a standard circular ROB model.  Section ~\ref{sec:circ} discusses its implementation.  We further model three optimizations for re-order buffers in an attempt to improve overall power usage: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch).  Each of these optimized models are built on top of the underlying circular ROB model to provide as close a behavioral and power comparison as possible.  Each of the optimizations, as well as the standard circular reorder buffer, are described in detail below.
\subsubsection{Circular ROB}
\label{sec:circ}
The circular ROB is the basic current implementation of the reorder buffer. It is implemented as a circular FIFO buffer, and the beginning and end of the used region are tracked by the head and tail pointers, respectively. When an instruction is issued, space is reserved for the instruction's result at the tail pointer, and the tail pointer is incremented. When the instruction completes, the result is inserted into the reorder buffer at the reserved space. The result is finally committed to the architecture register file when the head pointer passes through it; each cycle, results are committed from the head pointer and the head pointer is incremented repeatedly, until either an unfinished instruction result slot is reached or all write ports are used up.

The circular ROB is considered full when the tail pointer reaches the head pointer, as the head pointer denotes the start of the in-use region and no more space can be reserved from the tail pointer. Conversely, then buffer is empty when the head pointer catches up to the tail pointer, as every instruction in the buffer has now been committed to the ARF.

\subsubsection{Distributed (Cluster based) ROB}

\subsubsection{Dynamically Sized ROB}
The dynamically sized ROB is an optimization on the circular ROB where the buffer is divided into partitions, and those partitions deemed unnecessary are powered off to conserve energy. Our implementation of the basic dynamic ROB is behaviorally the same as that detailed in \cite{kucuk2}.

Downsizing the ROB is based on a periodic sampling of the buffer state. A number of cycles, {\it update\_period}, is defined, and every {\it update\_period} cycles the size of the ROB is updated. Every {\it sample\_period} cycles, the number of active entries is recorded, and the quantity {\it active\_size} is recorded as the average of these samples between each {\it update\_period}. The current maximum capacity of the reorder buffer is defined as the {\it current\_size}, and the size of a single ROB partition is defined as {\it partition\_size}. After each {\it update\_period} cycles, the ROB checks how full the buffer is, and if {\it current\_size} - {\it active\_size} > {\it partition\_size}, then a single partition is turned off.

Increasing the size of the ROB is dependent on how often the buffer fills up. A value, defined as the {\it overflow\_count} is incremented each cycle that instruction issue is stalled because the ROB is full. An {\it overflow\_threshold} is defined, and if during an {\it update\_period}, {\it overflow\_count} > {\it overflow\_threshold}, then a partition is turned on.

A conservative approach of deallocating strictly from one end of the buffer, and allocating strictly towards that end is followed. For example, if the ROB consisted of four partitions, downsizing the buffer would turn off the fourth, and then the third. At this point, if we were to increase the size of the buffer, the third partition would be turned on first, and the fourth partition next.

When the ROB's size is updated, partitions are not immediately turned off, since the partition could coincide with the active region of the buffer. If deallocating, this would disable a part of the active region, making instructions in the region impossible to complete. If allocating, this would insert an empty segment right into the middle of the FIFO queue, logically breaking the buffer. It is only safe to turn a partition off/on when the active region is not wrapped around (that is, the head pointer is less than the tail pointer), and the tail pointer comes before the first index into the partition to turn on/off. \cite{kucuk3} details a worst case scenario of deallocating, where the active region is wrapped (head is greater than tail), and both pointers lie in the region to be turned off. In this case, the head pointer must first clear the entries already in the buffer. Since the tail pointer is already in the partition, it will continue to write into the partition, and the head pointer must come around a second time to clear out these new entries. Only then can the partition be safely deallocated, after the head pointer has wrapped around to zero twice.

\subsubsection{Optimized Dynamic ROB}
We implement an optimized version of the dynamic ROB detailed in \cite{kucuk2} and \cite{kucuk3}, opting to be more aggressive with turning off partitions in order to save power, and to improve the partition allocation/deallocation scheme.

When downsizing the regular dynamic ROB, a single partition is turned off if {\it current\_size} - {\it active\_size} > {\it partition\_size}. In an effort to minimize the buffer size and thus the power consumed, we choose to downsize by

\begin{center}
$\lfloor \frac{current\_size - active\_size}{partition\_size} \rfloor$
\end{center}

Thus if much less than the capacity of the buffer is active, several partitions may be turned off at once.

To more efficiently resize the buffer, we found that choosing the first valid partition immediately after the one housing the tail pointer is optimal. For example, if we are allocating a partition in a four partition buffer, and the tail pointer resides in partition 3, we first check if partition 4 can be turned on, then partition 1, then partition 2.

In allocating the buffer, the reasoning is as follows. We aim to enable a disabled partition. It is impossible for the head pointer to lie in the disabled partition, thus there is no need to consider the head pointer. To maintain the logical consistency of the buffer, we cannot enable a partition that would divide the active region of the buffer. Thus, we must allocate a partition that comes after the tail pointer, and if we wish the buffer size to grow as soon as possible, the partition to be allocated must be the first disabled partition after the tail pointer.

In deallocating the buffer, turning off the first partition after the tail pointer is also optimal. Being after the tail pointer, once we mark the partition for disabling we can prevent the tail pointer from writing into it. This prevents the worst case scenario where the head pointer must wrap around 0 twice to disable a partition. The next enabled partition immediately after the tail pointer is also the one the active region will leave first (if the region is in the partition; if not the partition is already empty and ready to deactivate), thus it will empty out first. If we are to maximize the power savings, we must disable the first partition that is available to be disabled, thus the first active partition after the tail is the optimal choice.

One last optmization we implemented was just-in-time allocation of partitions. Normally, when the tail pointer increments, if a partition is not active the tail pointer simply skips it. In our optimization, before skipping the tail pointer will check if the disabled partition is marked to be enabled. If so, instead of skipping over it, the partition will be turned on and the tail pointer will write into it. This ensures that each partition to be turned on is enabled at the last possible moment, saving the energy that would otherwise be consumed by an empty partition waiting to be written into.

\subsubsection{Reduced Output Port ROB (Retention Latch)}

\subsection{Power Tabulation Implementation}
\section{Results}
hi
\subsection{subsection titl}
sub section.. with table!
\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

@@

%TODO add in bib
\begin{thebibliography}{------}
\bibitem{graphite}
MIT Graphite Simulator
https://github.com/mit-carbon/Graphite/wiki
\bibitem{kucuk}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Low-complexity reorder buffer architecture. In Proceedings
of the 16th international conference on Supercomputing (ICS '02). ACM, New York, NY, USA, 57-66. DOI=10.
1145/514191.514202 http://doi.acm.org/10.1145/514191.514202
\bibitem{kucuk2}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Energy-Efficient Design of the Reorder Buffer. In Proceedings of the International Workshop on Power and Timing Modeling, Optimization, and Simulation, September 2002.
\bibitem{kucuk3}
Gorhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2001. Reducing Power Requirements of Instruction Scheduling Through Dynamic Allocation of Multiple Datapath Resources. In Proceedings of the International Symposium on Microarchitecture, December 2001.
\bibitem{rupnow}
Kyle Rupnow, Arun Rodrigues, Keith Underwood, and Katherine Compton. 2006. Scientific applications vs. SPEC-FP: a comparison of program behavior. In Proceedings of the 20th annual international conference on Supercomputing (ICS '06). ACM, New York, NY, USA, 66-74. DOI=10.1145/1183401.1183413 http://doi.acm.org/10.1145/1183401.1183413
\end{thebibliography}
\end{document}
